---
title: "R Notebook"
output:
    html_document:
      code_download: true    
      theme: cosmo
      toc: true
      toc_float: true
      highlight: tango
      number_sections: true
---

To go back to the Home page, click [here](https://quentin-duchemin.github.io/ENPC-SDA/).

The data can be found [here](https://www.kaggle.com/uciml/breast-cancer-wisconsin-data/version/2).

Attribute Information:

1) ID number
2) Diagnosis (M = malignant, B = benign)
3-32)

Ten real-valued features are computed for each cell nucleus:

a) radius (mean of distances from center to points on the perimeter)
b) texture (standard deviation of gray-scale values)
c) perimeter
d) area
e) smoothness (local variation in radius lengths)
f) compactness (perimeter^2 / area - 1.0)
g) concavity (severity of concave portions of the contour)
h) concave points (number of concave portions of the contour)
i) symmetry
j) fractal dimension ("coastline approximation" - 1)

The mean, standard error and "worst" or largest (mean of the three
largest values) of these features were computed for each image,
resulting in 30 features. For instance, field 3 is Mean Radius, field
13 is Radius SE, field 23 is Worst Radius.

All feature values are coded with four significant digits.

Missing attribute values: none

Class distribution: 357 benign, 212 malignant




```{r}
library("tidyverse")
data <- read.csv('../../data/breast-cancer.csv');
data <- data.frame(data);
```

```{r}
features <- colnames(data)[3:32];
df_benign <- data[data$diagnosis=="B",features];
df_malignant <- data[data$diagnosis=="M",features];
```


```{r}
library(ggpubr)
library(ggplot2)
library(purrr)

myplot <- function(feature){
            qq.out <- qqplot(x=df_benign[[feature]], y=df_malignant[[feature]], plot.it=FALSE);
            qq.out <- as.data.frame(qq.out);
  
            # Set the x and y limits
            xylim <- range( c(qq.out$x, qq.out$y) );
            
            ggplot(qq.out, aes( x= x, y = y)) + 
                 geom_point() + 
                 geom_abline( intercept=0, slope=1) +
                 coord_fixed(ratio = 1, xlim=xylim, ylim = xylim) +
                 xlab("Benign") + ylab("Malignant") + labs(title=substr(feature,1,nchar(feature)-5))
}
plot_list <- map(features[1:4], myplot)
ggarrange(plotlist = plot_list, heights = c(10,10), nrow = 2,ncol = 2)

plot_list <- map(features[5:10], myplot)
ggarrange(plotlist = plot_list, heights = c(10,10), nrow = 3,ncol = 2)
```


```{r message=FALSE, warning=FALSE, include=FALSE}
pvalues <- vector(length=length(features));
names(pvalues) <- features;
for (feature in features){
    ksTest <- ks.test(df_benign[[feature]], df_malignant[[feature]]);
    print(colnames(ksTest));
    pvalues[feature] <- ksTest$p.value;
}
names(pvalues) <- features;
pvalues;
```


