---
title: "Logistic Regression with the SARS Cov 2 dataset"
output:
    html_document:
      code_download: true    
      theme: cosmo
      toc: true
      toc_float: true
      highlight: tango
      number_sections: true
---

To go back to the Home page, click [here](https://quentin-duchemin.github.io/ENPC-SDA/).

<br></br>

This notebook was inspired by a [post](https://towardsdatascience.com/predicting-the-probability-of-sars-cov-2-result-using-multiple-logistic-regressing-in-r-and-python-874cf9185d62) from the website [TowardsDataScience](https://towardsdatascience.com). The dataset can be found [here](https://www.kaggle.com/einsteindata4u/covid19). 


Our goal is twofold:
- We want to build a predictive model able to classify a patient as infected by the covid 19 or not based on his/her blood's data.
- We also want to identify diseases that have symptoms similar to the Covid 19. In that way, we can recommend to medical centers to treat separately people infected by these diseases. This would also allow to avoid to build dataset with false instances (i.e. with false discoveries).


# Data import and pre-processing

```{r include=FALSE}
library('DescTools')
library("tidyverse")
library('corrplot')
library('boot') # to use cv.glm for Cross-Validation
library('ROCR') # to use prediction and performance functions for ROC curves
library('ggplot2') # to use the function ggplot
library('gridExtra') # to use the command griad.arange to show multiple plots in the same window
library('lmtest') # to use the function lrtest for likelihood ratio tests 
```

```{r}
data <- read.csv('../../data/covid.csv');
data <- data.frame(data);
data$SARS.Cov.2.exam.result = 1*(data$SARS.Cov.2.exam.result == "positive")
```

The dataset contains many NaN values. This issue of missing values must be corrected before moving forward.

```{r}
dfDiseases = data[,append(seq(22,39),c(2,3))];
dfDiseases = dfDiseases[,-7];
dfDiseases = na.omit(dfDiseases);
dfPrediction  = data[,append(seq(7,20),c(2,3,40,42))];
dfPrediction = na.omit(dfPrediction);
```


# Prediction of Infection by the SARS Cov 2
## Data Exploration

### Study of Correlations 

```{r}
# attaching the final dataset
attach(dfPrediction)
# Correlation between variables
dfPrediction.corr = cor(dfPrediction)
dfPrediction.corr[,2]
# Correlation plot
corrplot(dfPrediction.corr, method = "square", type = "lower", tl.cex = 0.5);
```

We can take a first glance at the dataset and immediately identify some variables that are highly correlated with the target value (i.e. the variable SARS.Cov.2.exam.result).


### Usefulness of predictors


A first visual approach to identify the most relevant variables to predict infections by the SARS Cov 2 consists in fitting a logistic model with only one variable and to plot the probability of being infected by the Covid with respect to the value of this predictor. Let us illustrate this method for some features.


```{r}
# Plotting Separate probability graphs
plotting.function <- function(var,variableORrank){
  dfPrediction.sep.function = paste("SARS.Cov.2.exam.result", "~", as.character(var))
  dfPrediction.sep.glm = glm(as.formula(dfPrediction.sep.function), data = dfPrediction , family = binomial)
  cv.glm(dfPrediction,dfPrediction.sep.glm,K=10)$delta[1]
  
  dfPrediction.predicted.data <- data.frame(
    probability.of.SARS=dfPrediction.sep.glm$fitted.values,
    variable=dfPrediction[,as.character(var)])
  
  dfPrediction.predicted.data <- dfPrediction.predicted.data[
    order(dfPrediction.predicted.data$variable, decreasing=FALSE),]
  
  dfPrediction.predicted.data$rank <- 1:nrow(dfPrediction.predicted.data)
  
  ggplot(data=dfPrediction.predicted.data, aes(x= variable, y=probability.of.SARS)) +
    geom_point(aes(color=variable), size=3) +
    xlab(as.character(var)) +
    ylab("Probability of having SARS CoV-2") +
    scale_colour_gradient(low = "darkgreen", high = "darkred", na.value = NA) +
    ggtitle(coef(summary(dfPrediction.sep.glm))[,'Pr(>|z|)'])
}
plotfun1 = plotting.function("Platelets")
plotfun2 = plotting.function("Monocytes")
plotfun3 = plotting.function("Hemoglobin")
plotfun4 = plotting.function("Red.blood.cell.distribution.width..RDW.")
plotfun5 = plotting.function("Red.blood.Cells")
grid.arrange(plotfun1, plotfun2, plotfun3, plotfun4, plotfun5, ncol=3 , nrow = 2)
```

Except for platelets, all predictors have a positive correlation on the probability of having SARS CoV-2. If this method is a good way to grasp a first intuition, it remains qualitative and scaling aspects can be confusing. Let us now consider a more rigorous approach to identify relevant predictors by using the Likelihood Ratio Test.



```{r}
# Plotting Separate probability graphs
pvalue_LRTest <- function(var){
  globalnull <- glm(SARS.Cov.2.exam.result  ~ 1, data=dfPrediction);
  alternative <- glm(paste("SARS.Cov.2.exam.result", "~", as.character(var)),data=dfPrediction);
  res_lrtest <- lrtest(globalnull, alternative);
  res_lrtest$`Pr(>Chisq)`[2];
}
names_features <- colnames(dfPrediction);
names_features <- names_features[names_features != 'SARS.Cov.2.exam.result'];
pvalues <- sapply(names_features, pvalue_LRTest);
names(pvalues) <- names_features;
pvalues[order(pvalues, decreasing=FALSE)];
```



Based of the results provided by Likelihood ratio tests, we only consider the predictors leading to a p-value at most equal to 0.05. So for the analysis, we will use Leukocytes, Platelets, Monocytes, Eosinophils, Patient.Age.Quantile, Proteina.C.reativa.mg.dL, Basophils, Hemoglobin and Red.blood.Cells as predictors.


## Prediction with Logistic Regression


Let’s predict the error on our test dataset. 

```{r}
# Dividing Train/Test data with 80% training dataset
sample_size <- floor(0.8 * nrow(dfPrediction))
train_ind <- sample(nrow(dfPrediction), size = sample_size)
dfPrediction.train <- as.data.frame(dfPrediction[train_ind,])
dfPrediction.test <- as.data.frame(dfPrediction[-train_ind,])
# Logistic regression considering all the variables on the target variable SARS_COV2_Result
dfPrediction.function = paste("SARS.Cov.2.exam.result", "~", "Leukocytes + Platelets + Monocytes + Eosinophils + Patient.age.quantile + Proteina.C.reativa.mg.dL + Basophils + Hemoglobin + Red.blood.Cells")
dfPrediction.glm = glm(as.formula(dfPrediction.function), data = dfPrediction.train , family = binomial)
summary(dfPrediction.glm)
# 10 fold cross-validation to verify the model
cv.glm(dfPrediction.train,dfPrediction.glm,K=10)$delta[1]
# Predicting on test data based on training set
dfPrediction.glm.predict <- predict(dfPrediction.glm,dfPrediction.test,type = "response")
summary(dfPrediction.glm.predict)
tapply(dfPrediction.glm.predict, dfPrediction.test$SARS.Cov.2.exam.result, mean)
# Confusion matrix for threshold of 1%
dfPrediction.confusion = table(dfPrediction.test$SARS.Cov.2.exam.result, dfPrediction.glm.predict > 0.01)
rownames(dfPrediction.confusion) <- c("Predicted FALSE","Predicted TRUE");
print(dfPrediction.confusion);
# False negative error rate (Type II error)
dfPrediction.type2error = dfPrediction.confusion[1,1]/ (dfPrediction.confusion[1,1]+dfPrediction.confusion[2,2])
print(paste("The proportion of errors of Type II is ",as.character(dfPrediction.type2error)));
```

We plot below the ROC curve for our model.

```{r}
# Plotting ROCR curve
dfPrediction.ROCRpred = prediction(dfPrediction.glm.predict, dfPrediction.test$SARS.Cov.2.exam.result)
dfPrediction.ROCRperf = performance(dfPrediction.ROCRpred, "tpr", "fpr")
plot(dfPrediction.ROCRperf, colorize=TRUE, print.cutoffs.at=seq(0,1,by=0.1), text.adj=c(-0.2,1.7))
```

We plot the final results.

```{r}
# Creating a dataframe with variables and predicted values of SARS results
dfPrediction.predict.dataframe <- data.frame(
  probability.of.having.SARS=dfPrediction.glm$fitted.values,
  Leukocytes=dfPrediction.train$Leukocytes,
  Patient.age.quantile = dfPrediction.train$Patient.age.quantile,
  Eosinophils = dfPrediction.train$Eosinophils,
  Red.blood.cell.distribution.width..RDW. = dfPrediction.train$Red.blood.cell.distribution.width..RDW.,
  Platelets = dfPrediction.train$Platelets,
  Proteina.C.reativa.mg.dL = dfPrediction.train$Proteina.C.reativa.mg.dL)
plot1 = ggplot(data=dfPrediction.predict.dataframe, aes(x=Patient.age.quantile, y=probability.of.having.SARS)) +
  geom_point(aes(color=Patient.age.quantile), size=4)
plot2 = ggplot(data=dfPrediction.predict.dataframe, aes(x=Leukocytes, y=probability.of.having.SARS)) +
  geom_point(aes(color=Leukocytes), size=4)
plot3 = ggplot(data=dfPrediction.predict.dataframe, aes(x=Red.blood.cell.distribution.width..RDW., y=probability.of.having.SARS)) +
  geom_point(aes(color=Red.blood.cell.distribution.width..RDW.), size=4)
plot4 = ggplot(data=dfPrediction.predict.dataframe, aes(x=Eosinophils, y=probability.of.having.SARS)) +
  geom_point(aes(color=Eosinophils), size=4)
plot5 = ggplot(data=dfPrediction.predict.dataframe, aes(x=Platelets, y=probability.of.having.SARS)) +
  geom_point(aes(color=Platelets), size=4)
plot6 = ggplot(data=dfPrediction.predict.dataframe, aes(x=Proteina.C.reativa.mg.dL, y=probability.of.having.SARS)) +
  geom_point(aes(color=Proteina.C.reativa.mg.dL), size=4)
# Plotting the values
grid.arrange(plot1, plot2, plot3, plot4, plot5, plot6, ncol=3 , nrow = 2)
```


# Disease dataset

## Correlations

We have to skip Parainfluenza 2 as there are no cases for it.

```{r}
dfDiseases <- dfDiseases[,colnames(dfDiseases) != 'Parainfluenza.2'];
```

Now let’s plot the CramerV’s correlation plot to check the correlation between categorical variables.


```{r}
# CramerV Correlation to check for any correlation between catagorical variables
# Except Patient's age, all other variables are catagorical(binary)
dfDiseases.corr  = PairApply(dfDiseases[,names(dfDiseases) != "Patient.age.quantile"], CramerV, symmetric = TRUE)
# Displaying correlation with variable SARS_COV2_Result
dfDiseases.corr[,2]
# Correlation plot
corrplot(dfDiseases.corr , method = "square", type = "lower", number.cex=0.5)
```

Rhinovirus_OR_Enterovirus seems to be the predictor the most correlated with our target (with the correlation value of 0.1517). In the next section, we conduct a statistical test to identify if this correlation is statically significant.

## Likelihood Ratio Test

We perform Likelihood Ratio Tests to identify diseases that show a significant correlation with the infection status of patient by the SARS Cov 2.

```{r}
pvalue_LRTest <- function(var){
  globalnull <- glm(SARS.Cov.2.exam.result  ~ 1, data=dfDiseases);
  alternative <- glm(paste("SARS.Cov.2.exam.result", "~", as.character(var)),data=dfDiseases);
  res_lrtest <- lrtest(globalnull, alternative);
  res_lrtest$`Pr(>Chisq)`[2];
}
names_features <- colnames(dfDiseases);
names_features <- names_features[ !(names_features %in% c('SARS.Cov.2.exam.result','Patient.age.quantile'))];
pvalues <- sapply(names_features, pvalue_LRTest);
names(pvalues) <- names_features;
pvalues[order(pvalues, decreasing=FALSE)];
```
Our intuition of the previous section is confirmed by the Likelihood Ratio Test. The symptoms of Rhinovirus are close to the SARS CoV-2 virus and mostly found in the nose (Rhinovirus). Therefore, the probability of having SARS decreases as Rhinovirus is detected. 

Hence hospitals should distinguish between a person with Rhinovirus and a person with COVID-19. Separating patients with Rhinovirus and Covid-19 could help to relieve covid medical groups.
